Lower tax brackets and employment levels, and people living in populous areas are more likely to vote for Biden

### 2024-12-04 paper/paper.qmd
appendix, discussion, fixes on prose

---
title: "Younger age groups, women, and lower levels of education are more likely to vote Democractic"
subtitle: "A logistic regression analysis of demographics in the 2020 US Presidential Election"
author: 
  - Victor Ma
thanks: "Code and data are available at: https://github.com/bestmustard/political_support"
date: 12/3/2024
date-format: long
abstract: "In this study, I investigate demographic patterns which influence political support for the US 2020 election, specifically focusing on whether demographics of people tend to vote for Joe Biden or Donald Trump. I use a logistic regression model with the variables age, gender, and education to predict which categories in each demographic have the highest chance of voting for Joe Biden. I find that the model accurately predicts the trend of younger people, women, and those with the lowest level of education being more likely to support Joe Biden."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72

@@

# Introduction

The 2020 United States presidential election was a landmark event that encapsulated a confluence of crises and societal shifts. Beyond the COVID-19 pandemic, which reshaped healthcare priorities and exposed systemic vulnerabilities, the election underscored debates about economic inequality, particularly as millions faced unemployment during the economic downturn [@Blustein2020; @BLS2020]. Climate change emerged as a defining issue, with voters increasingly prioritizing environmental policies in response to intensifying natural disasters [@IPCC2020].

Demographic shifts also played a critical role. The increasing influence of younger, more diverse voters highlighted generational and racial divides in political priorities [@Frey2020]. Meanwhile, changes in suburban voting patterns, particularly among college-educated women, illustrated the evolving dynamics of key swing areas [@Cook2020].

Another major factor was the unprecedented role of mail-in voting, which became a focal point for controversy amidst the pandemic and claims of election fraud. This shift in voting methods not only increased accessibility for millions but also fueled political and legal battles over election integrity [@NCSL2020]. Finally, the election was a litmus test for the resilience of democratic institutions in the face of misinformation, political polarization, and challenges to the legitimacy of results [@BenGurion2020; @Sunstein2020].

Analyzing how living environments, employment status, and income levels correlate with political alignment in 2020 offers deep insights into societal trends. Research shows that urban residents tend to favor liberal candidates, reflecting the diversity and progressive social policies often prioritized in cities [@bishop2019bigsort]. Conversely, rural voters strongly supported Donald Trump, consistent with longstanding trends emphasizing traditional values and conservative policies [@cramer2016politics].

Employment status also played a key role. Data from the Cooperative Congressional Election Study (CCES, 2020) revealed that students and retirees, often focused on issues like education and healthcare, leaned more toward Joe Biden, while full-time workers, prioritizing economic stability, showed mixed preferences based on industry and region [@cces2020].

Income disparities further highlighted political divides. Lower-income groups, grappling with economic inequality, largely supported Biden due to his policies on healthcare and social safety nets, as noted in Pew Research’s 2020 voter analysis. Higher-income voters, on the other hand, leaned Republican, particularly in rural and suburban settings, reflecting preferences for tax policies that align with their financial interests [@pew2020voteranalysis].

These findings display the imapct of personal circumstances and political preferences, mapping not only the political landscape of 2020 but also providing a deeper understanding of the evolving priorities shaping voter behavior.

In this study, I use a logistic regression model to predict which candidate—Joe Biden or Donald Trump—was more likely to receive votes based on factors such as living environment, employment status, and income levels. Logistic regression is particularly suited for this analysis as it models binary outcomes, allowing us to estimate the likelihood of a voter aligning with one candidate over the other. By focusing on these explanatory variables, the study provides valuable insights into how socioeconomic and regional demographics influenced voter preferences during one of the most polarized elections in modern history.

The primary dataset for this study is the 2020 Cooperative Congressional Election Study (CCES), a stratified sample survey administered by YouGov and sourced from the Harvard Dataverse [@cces2020].

Complementary insights are drawn from Pew Research Center’s 2020 voter analysis, which contextualizes demographic trends across key groups, and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which provides aggregate statistics on voter turnout and demographic characteristics [@pew2020; @CPS2020]. These additional datasets validate findings from the CCES and enhance the robustness of our conclusions.

My report is structured into five main sections and includes an appendix. After the introduction, the first section describes the CCES 2020 dataset, highlighting key attributes and presenting graphs of demographic distributions. The second section details the logistic regression model, its rationale, and preliminary findings. The third section analyzes how living environment, employment status, and income influenced election outcomes using visualizations and numerical results. The fourth section discusses the implications of the findings, potential weaknesses, and future research directions.

The appendix provides an in-depth exploration of survey methodologies, focusing on sampling and observational data related to the CCES 2020 dataset. It includes detailed discussions of idealized methodologies, literature comparisons, and simulations to illustrate sampling variability and bias, expanding on points in the Data section.

This analysis was conducted in R, utilizing several packages to streamline data handling, visualization, and modeling. Key tools include tidyverse for data manipulation and visualization, dataverse for accessing the CCES dataset, knitr for generating the report, modelsummary for interpreting models, and rstanarm for Bayesian regression analysis [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr]. Additionally, ggplot visualizations and elements of the "Discussion" section were developed with the assistance of ChatGPT4 @chatgpt.
\newpage

# Data
The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse. The 2020 iteration surveyed 61,000 respondents, capturing a broad spectrum of U.S. political opinions. The release includes detailed documentation and questionnaires, ensuring transparency and replicability. Vote validation was conducted by Catalist, a large-scale organization maintaining data on over 240 million U.S. individuals [@Catalist2017].

@@

Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[], [@ces2020].

@@

### Data Collection Phases
The CCES collected data in two waves, immediately before and after the 2020 election. This helps capture the shifts in voter attitudes and preferences during the critical final stages of the election campaign [@ces2020]. Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture as broad of a perspective as the CES survey [@Pew2020], [@Gallup2020].

@@

## Variables of Interest
The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that underscores their influence on voting behavior.

@@

However, this trend is nuanced by education levels and geographic factors. High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].

@@

In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
These patterns underscore the nuanced relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.

@@

The @fig-heatmap-employment-living provides insights into the distribution of employment statuses across different living environments:

@@

@@

```
@fig-incomevotes shows that voting for Biden does not seem to have a correlation with voting for Biden.

@@

The model is implemented using R, leveraging the rstanarm package for Bayesian logistic regression. The data was preprocessed using the tidyverse package. Model outputs and intermediate steps are saved in .rds format for reproducibility [@citeR, @Rstanarm]. The model was implemented using the stan_glm() function, which applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors. The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.

@@

Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013]. Employment, income, and living environment are socio-economic factors that significantly influence voting preferences. These variables are categorical, making logistic regression suitable for analyzing their relationship with a binary outcome. The model avoids unnecessary complexity by focusing on these key variables while maintaining interpretability. Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without unduly constraining the model [@Gelman2008].

@@

Underlying assumptions include the linearity of predictors on the log-odds scale, independence of observations, and the assumption that predictors adequately capture variability in the data. Limitations include the categorical nature of the predictors, which may lose finer granularity, and the potential omission of confounders such as race or education level. The model may be less appropriate in cases where relationships are highly non-linear or involve strong interaction effects. Alternatives considered include random forests, which are more flexible but lack interpretability, and multinomial logistic regression, which was unnecessary for this binary outcome. Logistic regression was chosen for its balance of simplicity, interpretability, and compatibility with the data structure.

The validation process includes a train-test split to ensure generalization to new data, ROC Curve and AUC to demonstrate the model's discrimination ability with a moderate AUC value, posterior predictive checks to validate that simulated outcomes align well with observed data, and residual analysis to confirm model assumptions and identify any misfit. These steps confirm the robustness and reliability of the logistic regression model for political preferences.

@@

The analysis reveals clear patterns in the relationship between socio-economic predictors and voting preferences in the 2020 U.S. Presidential Election. Employment, income, and living environment significantly influence political alignment. The coefficients in @tbl-logit_summary display the impact of each predictor on the likelihood of voting for Joe Biden versus Donald Trump.

```{r}
#| label: tbl-model_summary
#| tbl-cap: "Coefficient Summary of the Logistic Regression Model"
#| echo: false
#| warning: false
#| message: false

@@

mcmc_intervals(posterior, prob = 0.9) +
  labs(
    title = "Coefficient Estimates",
    subtitle = "90% Credible Intervals",
    x = "Coefficient Value",
    y = "Predictors"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```
The visualization in @fig-coefficient_intervals shows the 90% credible intervals for each predictor. Predictors with intervals that do not overlap zero indicate significant effects on voting preferences. For example, Living Environment: Rural exhibits a significant negative association with voting for Biden, while Income: <10k shows a strong positive effect.

@@

```{r}
#| label: fig-income-overlay
#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Income Level"
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-employment-overlay
#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Employment"
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-living-overlay
#| fig-cap: "Overlay of Predicted Odds and Respondent Counts by Living Area"
#| echo: false
#| warning: false
#| message: false

#### Load model ####
political_preferences <- readRDS("../models/political_preferences.rds")

@@

```
@fig-living-overlay shows the model has completely different predictions for the living areas, when employment is set to "Full-time" and income is in the "200-500k" range. 
## Model Justification

@@

political_preferences <- readRDS(file ="../models/political_preferences.rds")
# Train-test split
set.seed(853)
train_indices <- sample(1:nrow(cleaned_data), 0.7 * nrow(cleaned_data))
train_data <- cleaned_data[train_indices, ]
test_data <- cleaned_data[-train_indices, ]

@@

In @fig-pp, The observed data $y$ is represented by the solid curve.
The replicated data $y_rep$ is represented by the shaded region, which encompasses the range of simulations based on the posterior distribution of the model parameters.

@@

This paper investigated the influence of socio-economic factors—employment status, income level, and living environment—on voting preferences during the 2020 U.S. Presidential Election. Utilizing the 2020 Cooperative Election Study (CCES) dataset, a logistic regression model was applied to predict support for Joe Biden versus Donald Trump. The model incorporated categorical predictors, regularized with Bayesian priors, to account for variability while ensuring interpretability. The model's predictors were carefully chosen to reflect significant socio-economic dimensions that shape voter behavior. Through robust statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.

The statistical approach balanced complexity and interpretability, leveraging Bayesian regularization to mitigate overfitting while allowing for nuanced insights. By employing a logistic regression model, this paper directly modeled the binary nature of the voting outcome while retaining a clear interpretation of how each predictor contributed to the likelihood of voting for a particular candidate. The inclusion of categorical variables like employment, income, and living environment provided an avenue to explore how socio-economic realities translate into political preferences.

## Insights

One key insight from this analysis is the continued significance of the urban-rural divide in American politics. Urban voters were far more likely to support Joe Biden, a trend consistent with existing literature on urban liberalism and rural conservatism [@Pew2020u]. Urban areas, characterized by higher population densities and greater economic and cultural diversity, often prioritize progressive policies such as environmental regulation, healthcare access, and social equity. Conversely, rural voters exhibited a marked preference for Donald Trump, reflecting a broader alignment with conservative policies, including lower taxation, gun rights, and traditional social values [@IyengarKrupenkin2018]. These findings underscore the stark political polarization between urban and rural constituencies, a division that has deepened over successive election cycles.

Another critical takeaway is the nuanced role of income in shaping political preferences. Lower-income groups were more likely to support Biden, aligning with the Democratic platform's emphasis on social welfare programs, affordable healthcare, and economic redistribution. Middle-income groups displayed more mixed preferences, often reflecting a balance between economic self-interest and policy priorities. Interestingly, the wealthiest demographic groups exhibited reduced support for Biden, suggesting that concerns over Democratic tax policies and fiscal regulations may outweigh social considerations in shaping their voting behavior [@Frank2004; @Gelman2009]. This U-shaped relationship between income and voting patterns highlights the complexity of economic factors in political decision-making.

@@

While this study provides valuable insights, several limitations warrant discussion. First, the logistic regression model assumes linearity in the log-odds of the predictors, which may oversimplify complex interactions between socio-economic factors and voting behavior. Non-linear relationships or interaction effects, such as how income and employment jointly influence voting preferences, may be overlooked in this framework.

The categorical representation of predictors, while simplifying interpretation, also introduces granularity constraints. For example, income categories like "50-100k" encompass a broad range of experiences that may differ substantially in their political implications. Similarly, the employment categories do not account for nuances such as industry type or job satisfaction, which could have significant effects on political alignment.

@@

Finally, the model does not account for regional variations within urban and rural categories. For instance, urban areas in different states may have distinct political leanings influenced by local economic conditions, cultural dynamics, or historical contexts. Ignoring these regional nuances may obscure important variations in the data.

@@

The findings of this paper open several avenues for future research. A key area for exploration is the temporal dimension of voting behavior. Incorporating longitudinal data could provide insights into how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time. Such an approach would allow researchers to examine causal relationships rather than mere associations.

Intersectionality is another promising direction. Investigating how combinations of factors—such as race, gender, and education—jointly influence voting behavior could yield a richer understanding of political alignment. For example, how do the political preferences of low-income individuals differ by gender or race? Addressing these questions would require a more granular dataset and advanced modeling techniques, such as hierarchical or multi-level models [@Highton2009].

@@

Finally, regional analyses could shed light on how local contexts mediate the relationship between socio-economic factors and voting behavior. For example, exploring how employment trends in manufacturing-heavy states compare to service-oriented urban centers could reveal additional layers of complexity in voter alignment. Incorporating regional economic data or policy differences could enhance the explanatory power of future models.

In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also underscores the need for more nuanced and dynamic approaches to understanding voter behavior. By addressing these limitations and pursuing the outlined future directions, researchers can build a more comprehensive and context-sensitive picture of the socio-political landscape.

\newpage

>>>>

---
title: "Lower tax brackets and employment levels, and people living in populous areas are more likely to vote for Biden"
author: 
  - Victor Ma
thanks: "Code and data are available at: https://github.com/bestmustard/political_support"
date: 12/3/2024
date-format: long
abstract: "In this study, I investigate demographic patterns that influence political support during the 2020 U.S. Presidential Election, focusing on whether socio-economic factors like employment status, income level, and living environment influence voters' likelihood to support Joe Biden or Donald Trump. I employ a logistic regression model to analyze these predictors and identify trends within each demographic category. The results show that urban residents, lower-income groups, and individuals in specific employment categories (e.g., students and retirees) are more likely to support Joe Biden. These findings align with broader trends in U.S. politics, where socio-economic conditions and regional contexts play pivotal roles in shaping political preferences."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72

@@

# Introduction
The 2020 United States presidential election highlighted key societal shifts amidst crises. Beyond the COVID-19 pandemic, which reshaped healthcare and exposed systemic vulnerabilities, the election underscored debates about economic inequality as unemployment surged [@Blustein2020; @BLS2020]. Climate change became a priority for many voters, driven by the growing frequency of natural disasters [@IPCC2020].

Demographic patterns also shaped the election. Younger, more diverse voters emphasized generational and racial divides in priorities [@Frey2020]. Suburban voting patterns, particularly among college-educated women, had shifts in key swing areas [@Cook2020]. The widespread adoption of mail-in voting during the pandemic increased accessibility but fueled political and legal debates over election integrity [@NCSL2020]. The election also tested democratic institutions against misinformation and polarization [@BenGurion2020; @Sunstein2020].

Living environments strongly influenced political alignment. Urban residents predominantly supported Joe Biden, reflecting progressive policy preferences often found in cities [@bishop2019bigsort]. Conversely, rural voters favored Donald Trump, consistent with support for conservative values [@cramer2016politics].

Employment status was another factor. Data from the Cooperative Congressional Election Study (CCES, 2020) indicated that students and retirees leaned toward Biden, likely due to policy interests in education and healthcare. Full-time workers exhibited mixed preferences influenced by industry and region [@cces2020].

Income disparities also defined political divides. Lower-income groups largely supported Biden, aligning with Democratic policies on healthcare and social safety nets, while higher-income voters leaned Republican, especially in rural and suburban settings, reflecting preferences for tax policies benefiting their financial interests [@pew2020voteranalysis].

This study uses a logistic regression model to predict voting preferences based on living environment, employment status, and income levels. Logistic regression is suited for binary outcomes, providing probabilities for voter alignment with each candidate.

The primary dataset is the 2020 Cooperative Congressional Election Study (CCES), a stratified survey conducted by YouGov and sourced from the Harvard Dataverse [@cces2020]. Supporting data includes Pew Research Center’s 2020 voter analysis and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which validate and contextualize findings [@pew2020; @CPS2020].

The report includes five sections and an appendix. After the introduction, the first section describes the CCES dataset, highlighting demographic attributes and distributions. The second section explains the logistic regression model and presents preliminary findings. The third section analyzes the influence of living environment, employment status, and income on election outcomes using visualizations and numerical results. The fourth section discusses the implications, limitations, and directions for future research.

The appendix explores survey methodologies, focusing on sampling and observational data in the CCES dataset. It compares methodologies, discusses sampling variability and bias, and uses simulations to expand on points in the Data section.

This analysis was done in R, with the packages tidyverse for data manipulation, dataverse for dataset access, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr]. Plots and graphs were created with ggplot, and portions of the discussion, data, and appendix were assisted by ChatGPT4 [@chatgpt].
\newpage

# Data
The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse. The 2020 iteration surveyed 61,000 respondents, capturing a broad range of U.S. political opinions. The release includes detailed documentation and questionnaires, ensuring transparency and replicability. Vote validation was conducted by Catalist, a large-scale organization maintaining data on over 240 million U.S. individuals [@Catalist2017].

@@

Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Edison2020], [@ces2020].

@@

### Data Collection Phases
The CCES collected data in two waves, immediately before and after the 2020 election. This helps capture the shifts in voter attitudes and preferences during the final stages of the election campaign [@ces2020]. Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture a longitudinal perspective similarly to the CES survey [@Pew2020].

@@

## Variables of Interest
The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that shows their influence on voting behavior.

@@

However, this trend is affected by education levels and geographic factors. High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].

@@

In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
These patterns underscore the relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.

@@

The @fig-heatmap-employment-living shows the distribution of employment statuses across different living environments:

@@

#### Generate income_summary ####
income_summary <- cleaned_data %>%
  group_by(income) %>%
  summarize(
    proportion_biden = mean(voted_for == "Biden", na.rm = TRUE),
    total_count = n()
  )

@@

@fig-incomevotes shows that income level does not seem to have a correlation with voting for Biden.

@@

The model is implemented using R, using the rstanarm package for Bayesian logistic regression. The data was preprocessed using the tidyverse package. Model outputs and intermediate steps are saved in .rds format for reproducibility [@citeR, @Rstanarm]. The model was implemented using the stan_glm() function, which applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors. The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.

@@

Employment, income, and living environment are socio-economic factors that significantly influence voting preferences. These variables are categorical, making logistic regression suitable for analyzing their relationship with a binary outcome. The model avoids unnecessary complexity by focusing on these key variables while maintaining interpretability. Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without constraining the model [@Gelman2008].

@@

Underlying assumptions include the linearity of predictors on the log-odds scale, independence of observations, and the assumption that predictors adequately capture variability in the data. Limitations include the categorical nature of the predictors, which may lose granularity, and the potential omission of confounders such as race or education level. The model may be less appropriate in cases where relationships are non-linear or involve strong interaction effects. Alternatives considered include random forests, which are more flexible but lack interpretability, and multinomial logistic regression, which was unnecessary for this binary outcome. Logistic regression was chosen for its balance of simplicity, interpretability, and compatibility with the data structure.

The validation process includes a train-test split to ensure generalization to new data, ROC Curve and AUC to demonstrate the model's discrimination ability with a moderate AUC value, posterior predictive checks to validate that simulated outcomes align well with observed data, and residual analysis to confirm model assumptions and identify any misfit. These steps confirm the reliability of the logistic regression model for this use case.

@@

The resuls of this analysis show that employment, income, and living environment significantly influence political alignment. The coefficients in @tbl-model_summary display the impact of each predictor on the likelihood of voting for Joe Biden versus Donald Trump.

```{r}
#| label: tbl-model_summary
#| tbl-cap: "Coefficient Summary of the Logistic Regression Model"
#| echo: false
#| warning: false
#| message: false
political_preferences <- readRDS(file ="../models/political_preferences.rds")

@@

mcmc_intervals(posterior, prob = 0.9) +
  labs(
    title = "Coefficient Estimates",
    subtitle = "90% Credible Intervals",
    x = "Coefficient Value",
    y = "Predictors"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```
The visualization in @fig-coefficient_intervals shows the 90% credible intervals for each predictor. Predictors with intervals that do not overlap zero indicate significant effects on voting preferences. For example, livingRural area exhibits a significant negative association with voting for Biden, while employmentStudent shows a strong positive effect.

@@

```{r out.height="20%"}
#| label: fig-income-overlay
#| fig-cap: "Model Prediction vs. Actual Votes by Income Level"
#| echo: false
#| warning: false
#| message: false

@@

```{r out.height="20%"}
#| label: fig-employment-overlay
#| fig-cap: "Model Prediction vs. Actual Votes by Employment"
#| echo: false
#| warning: false
#| message: false

@@

```{r out.height="20%"}
#| label: fig-living-overlay
#| fig-cap: "Model Prediction vs. Actual Votes by Living Area"
#| echo: false
#| warning: false
#| message: false

#### Load model ####

@@

```
@fig-living-overlay shows the model has completely different predictions for the living areas, when employment is set to "Full-time" and income is in the "200-500k" range.

## Model Evaluation

@@

# Train-test split
set.seed(853)
train_indices <- sample(1:nrow(cleaned_data), 0.7 * nrow(cleaned_data))
train_data <- cleaned_data[train_indices, ]
test_data <- cleaned_data[-train_indices, ]

@@

In @fig-pp, The observed data $y$ is represented by the solid curve.
The replicated data $y_rep$ is represented by the shaded region, which covers the range of simulations based on the posterior distribution of the model parameters.

@@

This paper investigated the influence of socio-economic factors—employment status, income level, and living environment—on voting preferences during the 2020 U.S. Presidential Election. Utilizing the 2020 Cooperative Election Study (CCES) dataset, a logistic regression model was applied to predict support for Joe Biden versus Donald Trump. The model incorporated categorical predictors, regularized with Bayesian priors, to account for variability while ensuring interpretability. The model's predictors were carefully chosen to reflect significant socio-economic dimensions that shape voter behavior. Through statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.

By employing a logistic regression model, this paper directly modeled the binary nature of the voting outcome while retaining a clear interpretation of how each predictor contributed to the likelihood of voting for a particular candidate. The inclusion of categorical variables like employment, income, and living environment provided an avenue to explore how socio-economic realities translate into political preferences.

## Takeaways

This analysis shows the continued significance of the urban-rural divide in American politics. Urban voters were far more likely to support Joe Biden, a trend consistent with existing literature on urban liberalism and rural conservatism [@Pew2020u]. Urban areas, characterized by higher population densities and greater economic and cultural diversity, often prioritize progressive policies such as environmental regulation, healthcare access, and social equity. Conversely, rural voters exhibited a marked preference for Donald Trump, reflecting a broader alignment with conservative policies, including lower taxation, gun rights, and traditional social values [@IyengarKrupenkin2018]. These findings underscore the stark political polarization between urban and rural constituencies, a division that has deepened over successive election cycles.

Another takeaway is the role of income in shaping political preferences. Lower-income groups were more likely to support Biden, aligning with the Democratic platform's emphasis on social welfare programs, affordable healthcare, and economic redistribution. Middle-income groups displayed more mixed preferences, often reflecting a balance between economic self-interest and policy priorities. Interestingly, the wealthiest demographic groups exhibited reduced support for Biden, suggesting that concerns over Democratic tax policies and fiscal regulations may outweigh social considerations in shaping their voting behavior [@Frank2004; @Gelman2009]. This U-shaped relationship between income and voting patterns highlights the complexity of economic factors in political decision-making.

@@

While the model was well-suited for the purpose of predicting binary outcomes, there were several limitations. First, the logistic regression model assumes linearity in the log-odds of the predictors, which may oversimplify complex interactions between socio-economic factors and voting behavior. Non-linear relationships or interaction effects, such as how income and employment jointly influence voting preferences, may be overlooked in this framework.

The categorical representation of predictors, while simplifying interpretation, also introduces granularity constraints. For example, income categories like "50-100k" cover a broad range of experiences that may differ substantially in their political implications. Similarly, the employment categories do not account for factors such as industry type or job satisfaction, which could have significant effects on political alignment.

@@

Finally, the model does not account for regional variations within urban and rural categories. For instance, urban areas in different states may have distinct political leanings influenced by local economic conditions, cultural dynamics, or historical contexts. Ignoring these regional differences may obscure important variations in the data.

@@

The findings of this paper open several avenues for future research. A key area for discussion is the role of timing in voting behavior. Incorporating longitudinal data could describe how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time. Such an approach would allow researchers to examine causal relationships rather than mere associations.

Intersectionality is another promising direction. Investigating how combinations of factors—such as race, gender, and education—jointly influence voting behavior could yield a richer understanding of political alignment. For example, how do the political preferences of low-income individuals differ by gender or race? Addressing these questions would require a more granular dataset and modeling techniques such as hierarchical or multi-level models [@Highton2009].

@@

Regional analyses are necessary to show how local contexts mediate the relationship between socio-economic factors and voting behavior. For example, employment trends in manufacturing-heavy states compared to service-oriented urban centers might cause differences in voter alignment. Incorporating regional economic data or policy differences could lead to mroe accurate explanations in future models.

In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also shows the need for more dynamic approaches to understanding voter behavior. By addressing these limitations and pursuing the outlined future directions, researchers can build a more comprehensive and context-sensitive picture of the socio-political landscape.

# Appendix: Survey and Sampling Methodology

Overview of Survey Design in the CCES 2020 Dataset

The 2020 Cooperative Election Study (CCES) is a large-scale, national survey designed to capture U.S. political behavior and attitudes. It employs a stratified sample, drawing from a diverse pool of respondents to achieve national representation The survey design is a collaborative effort between academic institutions and professional survey firms [@Ansolabehere2012].

The CCES includes two waves: a pre-election wave conducted in October 2020 and a post-election wave conducted in November 2020. This dual-wave structure allows researchers to capture shifts in voter sentiment and behavior during the election period. The survey uses online data collection, a method that benefits from scalability and cost-effectiveness.

## Sampling Techniques

### Stratified Sampling

Stratified sampling ensures that subgroups of the population, such as those defined by age, race, income, or geographic region, are adequately represented. For the CCES, stratification is based on demographic characteristics derived from the U.S. Census Bureau and voter registration databases. This method minimizes sampling error and allows for more precise estimates of population parameters [@Lohr2021].

### Sample Matching

Sample matching is a two-step process used to improve the representativeness of the survey sample. First, a target sample is generated based on demographic benchmarks. Next, survey respondents are selected from a larger pool to match the characteristics of the target sample. The CCES utilizes voter registration data to match respondents, enhancing the validity of findings [@Rivers2006].

### Weighting

Post-stratification weights are applied to the data to correct for any deviations between the survey sample and the target population. These weights adjust for factors such as response bias and demographic discrepancies. The weights are calculated using iterative proportional fitting, ensuring alignment with key benchmarks like age, gender, race, and geographic distribution [@Gelman2007].

## Observational Data and Challenges

### Strengths of Observational Surveys

Observational surveys like the CCES provide a snapshot of real-world voter behavior, capturing smaller details that experimental designs may overlook. They allow for large-scale data collection across diverse populations, which is important for studying complex phenomena like political alignment.

### Limitations

Despite their strengths, observational surveys face challenges such as:

1. Nonresponse Bias: Certain demographic groups may be less likely to participate, leading to underrepresentation.

2. Self-Reporting Bias: Respondents may misreport their voting behavior or demographic characteristics, often due to social desirability [@Ansolabehere2012].

3. Cross-Sectional Nature: The CCES captures data at specific points in time, limiting its ability to account for temporal changes in voter behavior.

## Comparisons with Other Surveys

### American National Election Studies (ANES)

The ANES employs a mixed-mode approach, combining face-to-face interviews with online surveys. While ANES provides rich contextual data, its smaller sample size limits the granularity of subgroup analysis compared to the CCES [@Berinsky2014].

### Pew Research Center Surveys

Pew Research specializes in tracking public opinion trends but focuses on shorter questionnaires. The CCES’s larger number of more specific questions is more suited for modeling complex relationships like those explored in this paper [@Pew2020].

## Methodological Best Practices

Based on the strengths and limitations of the CCES and other surveys, several best practices emerge:

1. Hybrid Sampling Methods: Combining stratified sampling with random digit dialing or online panel recruitment can improve representativeness.

2. Enhanced Weighting Techniques: Incorporating machine learning algorithms into weighting procedures could reduce bias from hard-to-reach populations.

3. Longitudinal Design: Adding a longitudinal component to surveys like the CCES would allow researchers to track changes in voter behavior over time.

Future iterations of surveys like the CCES could benefit from integrating administrative data (e.g., tax records or employment data) to validate self-reported responses. Additionally, adopting adaptive sampling techniques that dynamically adjust to underrepresented groups could further enhance data quality. Finally, the incorporation of qualitative components, such as open-ended responses, could provide richer context to quantitative findings.

By adhering to these methodological enhancements, surveys can accurately represent complex socio-political phenomena.
\newpage
